---
title: "Week 3 — Notebook · Richer models"
subtitle: "Gause competition, carrying capacity, parameter play"
format:
  html:
    toc: true
    css: ../../styles.scss
execute:
  echo: true
  warning: false
  message: false
---

## Learning goals

1. Compare *Paramecium caudatum* and *P. aurelia* alone vs together
2. Implement logistic growth and two-species competition updates
3. Change one parameter at a time and interpret biologically

## 1. Data first (Gause 1934, Fig. 22)

```{r}
pc <- read.csv("data/paramecium_caudatum_alone.csv")
pa <- read.csv("data/paramecium_aurelia_alone.csv")
mix <- read.csv("data/paramecium_competition.csv")

par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
plot(pc$day, pc$volume, type = "b", pch = 16,
     xlab = "Day", ylab = "Volume", main = "Caudatum alone")
plot(pa$day, pa$volume, type = "b", pch = 17,
     xlab = "Day", ylab = "Volume", main = "Aurelia alone")
plot(mix$day, mix$caudatum, type = "b", pch = 16,
     xlab = "Day", ylab = "Volume", main = "Mixture",
     ylim = range(c(mix$caudatum, mix$aurelia), na.rm = TRUE))
lines(mix$day, mix$aurelia, type = "b", pch = 17, lty = 2)
legend("topright", c("caudatum", "aurelia"), pch = c(16, 17), lty = 1:2, bty = "n", cex = 0.8)
```

**Q1.** In 2–3 sentences: what biological ingredients are missing from Week 2 predator–prey LV?

## 2. Logistic one-step update

```{r}
next_logistic <- function(N, r, K, dt) {
  dN <- r * N * (1 - N / K) * dt
  max(N + dN, 0)
}

next_logistic(N = 10, r = 0.8, K = 100, dt = 0.1)
```

**Q2.** If \(N\) is near \(K\), does the next step grow or shrink? Why?

## 3. Two-competitor update + loop

```{r}
next_comp <- function(N1, N2, r1, r2, K1, K2, a12, a21, dt) {
  d1 <- r1 * N1 * (1 - (N1 + a12 * N2) / K1) * dt
  d2 <- r2 * N2 * (1 - (N2 + a21 * N1) / K2) * dt
  c(N1 = max(N1 + d1, 0), N2 = max(N2 + d2, 0))
}

T <- 400
dt <- 0.05
N1 <- numeric(T + 1)
N2 <- numeric(T + 1)
N1[1] <- 20
N2[1] <- 20

r1 <- 1.0; r2 <- 1.0
K1 <- 150; K2 <- 150
a12 <- 1.2   # effect of species 2 on 1
a21 <- 0.5   # effect of species 1 on 2

for (t in 1:T) {
  out <- next_comp(N1[t], N2[t], r1, r2, K1, K2, a12, a21, dt)
  N1[t + 1] <- out["N1"]
  N2[t + 1] <- out["N2"]
}

time <- 0:T * dt
plot(time, N1, type = "l", lwd = 2, ylim = c(0, max(c(N1, N2))),
     xlab = "Time", ylab = "Abundance", main = "Competition simulation")
lines(time, N2, lwd = 2, lty = 2)
legend("topright", c("N1", "N2"), lty = 1:2, lwd = 2, bty = "n")
```

## 4. Experiment

Change **only** `a12` (or only `K1`). Re-run.

**Q3.** Did you get exclusion? Which species “won”? Link back to the mixture panel.

## Wrap-up

- Main quantitative result:
- Main biological interpretation:
