---
title: "Week 3 — Making Models Richer"
subtitle: "Extending Lotka–Volterra in code"
author: "Stephan Peischl"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    embed-resources: true
    transition: none
    controls: true
    controls-layout: edges
    controls-tutorial: true
    navigation-mode: linear
    hash: true
    menu:
      side: left
      use-text-content: true
    footer: "[Course outline](../../docs/course-outline.html) · [Home](../../index.html)"
execute:
  echo: false
  warning: false
  message: false
  freeze: true
---

## On this deck

**Week 3** of 14 · Ecology block

- [Course outline](../../docs/course-outline.html)
- [Home](../../index.html)
- Arrow keys / controls · **M** menu · **Esc** grid

## Part 1  
## Recap and goal

## Where we are

Last week: *Paramecium* + *Didinium* → discrete LV → function / vector / loop.

Today: **richer biology** with the same engine — motivated by another real Gause experiment.

## Learning goals today

1. Read a real **competition** contrast in data
2. Add carrying capacity \(K\) in code
3. Extend to **two competing species**
4. Change one parameter at a time and interpret

---

## Part 2  
## Bio opener: two Paramecia

## Not predator–prey this time

Same genus, shared resource:

- ***Paramecium caudatum***
- ***Paramecium aurelia***

Question: what happens **alone** vs **together**?

**Source:** Gause (1934) Fig. 22 — digitized in `gauseR` (`data/`).

## Alone: each grows then levels off

```{r}
#| fig-height: 3.8
#| fig-width: 8
pc <- read.csv("data/paramecium_caudatum_alone.csv")
pa <- read.csv("data/paramecium_aurelia_alone.csv")
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(pc$day, pc$volume, type = "b", pch = 16, lwd = 2,
     xlab = "Day", ylab = "Volume", main = "P. caudatum alone")
plot(pa$day, pa$volume, type = "b", pch = 17, lwd = 2,
     xlab = "Day", ylab = "Volume", main = "P. aurelia alone")
```

Both look **logistic** — a carrying capacity, not endless exponential growth.

## Together: competitive exclusion

```{r}
#| fig-height: 4
#| fig-width: 7
mix <- read.csv("data/paramecium_competition.csv")
ymax <- max(c(mix$caudatum, mix$aurelia), na.rm = TRUE)
plot(mix$day, mix$caudatum, type = "b", pch = 16, lwd = 2,
     xlab = "Day", ylab = "Volume",
     main = "Mixture (Gause 1934, Fig. 22)",
     ylim = c(0, ymax * 1.05))
lines(mix$day, mix$aurelia, type = "b", pch = 17, lwd = 2, lty = 2)
legend("topright", c("P. caudatum", "P. aurelia"),
       pch = c(16, 17), lty = 1:2, bty = "n")
```

## What jumps out?

- Alone: both persist near a plateau \(\approx K\)
- Together: **one declines** while the other holds on
- Classic message: competition can exclude a species that would do fine alone

## Think–pair–share (2 min)

1. Why isn’t Week 2’s predator–prey LV enough here?
2. What new ingredients do we need (\(K\)? two competitors?)
3. What would you change first in `next_state()`?

---

## Part 3  
## Critique last week’s simple LV

## Classic LV assumptions (too strong)

- Unlimited prey growth without predators
- No carrying capacity
- Only predator–prey (no competition)
- Deterministic (no noise) — later in the course

## Biological upgrades (menu)

1. Prey self-limitation (carrying capacity \(K\)) ← **today**
2. Two-species **competition** ← **today**
3. Predator saturation / refuges
4. Harvesting term

Each upgrade = a **small change in the update rule**.

---

## Part 4  
## Extension 1: carrying capacity

## Biology

Populations cannot grow forever even without enemies.

\[
\frac{dN}{dt} = rN\left(1 - \frac{N}{K}\right)
\]

Matches the **alone** curves in Fig. 22.

## With a predator (optional link to Week 2)

\[
\frac{dN}{dt} = rN\left(1 - \frac{N}{K}\right) - aNP
\]

## In discrete code

```r
dN <- (r * N * (1 - N / K) - a * N * P) * dt
```

Programming lesson:

> Change biology → change one expression inside the function.

## Live comparison

Run logistic alone vs LV with \(K\):

- Does \(N\) plateau near \(K\) when \(P = 0\)?
- How do cycles change when \(K\) is finite?

---

## Part 5  
## Extension 2: two competitors

## Biology (Lotka–Volterra competition)

Two species, shared limiting resource:

\[
\frac{dN_1}{dt} = r_1 N_1 \left(1 - \frac{N_1 + \alpha_{12} N_2}{K_1}\right)
\]
\[
\frac{dN_2}{dt} = r_2 N_2 \left(1 - \frac{N_2 + \alpha_{21} N_1}{K_2}\right)
\]

Goal: qualitatively recover **exclusion** like Gause Fig. 22.

## Sketch in code

```r
next_comp <- function(N1, N2, r1, r2, K1, K2, a12, a21, dt) {
  d1 <- r1 * N1 * (1 - (N1 + a12 * N2) / K1) * dt
  d2 <- r2 * N2 * (1 - (N2 + a21 * N1) / K2) * dt
  c(N1 = max(N1 + d1, 0), N2 = max(N2 + d2, 0))
}
```

Same Week 2 pattern — only **more components**.

## Store trajectories

- separate vectors, or
- a matrix (time × species)

Preview of later matrix thinking.

---

## Part 5  
## Parameter play as scientific method

## The habit we want

1. Change **one** thing
2. Predict what should happen
3. Simulate
4. Compare prediction vs result
5. Explain biologically

## What to explore today

In the notebook, start from Gause competition data, then:

- fit the *idea* of logistic growth with \(K\) (qualitative)
- simulate two-competitor LV; try to get exclusion
- vary \(r\), \(K\), or competition coefficients **one at a time**

## Plotting for insight

- alone vs mixture panels (data)
- simulated \(N_1(t), N_2(t)\) overlay
- (optional) \(N_1\) vs \(N_2\) phase plane

## Think-pair-share

If *P. aurelia* is the stronger competitor in mixture:

- Which coefficient would you increase first?
- How would monoculture plateaus constrain \(K_1, K_2\)?
- How is this different from Week 2’s *Didinium* story?

---

## Part 6  
## Notebook & wrap-up

## Notebook today

**Competition data → logistic / two-species model → parameter play**

1. Plot Gause Fig. 22 alone vs mixture (`data/`)
2. Working logistic + competition `next_state`
3. Students change one parameter; interpret biologically
4. Optional: add a predation term back (link Week 2)

## Bridge to programming course

| Programming | Biology today |
|---|---|
| Edit expressions in a function | add \(K\) / competition |
| Re-run scripts | alone vs mixture scenarios |
| Multiple related plots | communicate exclusion |
| Longer return vectors | two-species state |

## Exit ticket

1. What pattern in Fig. 22 requires a model richer than Week 2 LV?
2. In words: how does \(K\) change the update for one species alone?
3. What surprised you when you changed a competition parameter?

## Next week

- **Isle Royale** moose–wolf (and optional classic series)
- Model zoo: competition, epidemics, harvest…
- Notebook: **build your own model**

**Data sources:** Gause (1934) Fig. 22 via `gauseR` — see `data/README.md`.

---

## Navigation

| | |
|---|---|
| Overview | [Course outline](../../docs/course-outline.html) |
| Home | [Course home](../../index.html) |
| Previous | [← Week 2](../week-02/slides.html) |
| Next | [Week 4 →](../week-04/slides.html) |

Keyboard: ← → move slides · **M** menu · **Esc** slide grid · **F** fullscreen

