---
title: "Week 4 — Notebook · Your own model"
subtitle: "Isle Royale data → design → simulate"
format:
  html:
    toc: true
    css: ../../styles.scss
execute:
  echo: true
  warning: false
  message: false
---

## Learning goals

1. Inspect a real ecological time series qualitatively
2. Write verbal rules before code
3. Implement a small model with the Week 2–3 engine
4. Compare simulation patterns to data (no formal fitting)

## 1. Look at Isle Royale

```{r}
mw <- read.csv("data/isle_royale_moose_wolf.csv")
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(mw$year, mw$moose, type = "b", pch = 16,
     xlab = "Year", ylab = "Moose", main = "Isle Royale moose")
plot(mw$year, mw$wolf, type = "b", pch = 17,
     xlab = "Year", ylab = "Wolves", main = "Isle Royale wolves")
```

**Q1.** List three features of these series (trend, lag, crash, noise…).

Optional extras in `data/`: `huffaker_mites.csv`, `hudson_bay_lynx.csv`.

## 2. Design (words first)

Fill in:

1. State variables (2–3):
2. What increases each?
3. What decreases each?
4. One question you will ask the model:

## 3. Code scaffold

Adapt this predator–prey skeleton — or replace with competition / SIR / harvest.

```{r}
next_state <- function(N, P, r, a, b, m, K, dt) {
  dN <- (r * N * (1 - N / K) - a * N * P) * dt
  dP <- (b * a * N * P - m * P) * dt
  c(N = max(N + dN, 0), P = max(P + dP, 0))
}

T <- 800
dt <- 0.05
N <- numeric(T + 1)
P <- numeric(T + 1)
N[1] <- 40
P[1] <- 5

r <- 0.8; a <- 0.05; b <- 0.4; m <- 0.3; K <- 120

for (t in 1:T) {
  out <- next_state(N[t], P[t], r, a, b, m, K, dt)
  N[t + 1] <- out["N"]
  P[t + 1] <- out["P"]
}

time <- 0:T * dt
plot(time, N, type = "l", lwd = 2, xlab = "Time", ylab = "Abundance",
     main = "Your simulation", ylim = c(0, max(c(N, P))))
lines(time, P, lwd = 2, lty = 2)
legend("topright", c("Prey-like", "Predator-like"), lty = 1:2, lwd = 2, bty = "n")
```

## 4. Compare & reflect

**Q2.** Which qualitative features of Isle Royale does your model capture? Which does it miss?

**Q3.** What is the strongest assumption you made?

## Wrap-up

- Model question:
- One plot takeaway:
- Next experiment if you had more time:
