---
title: "Week 2 — Notebook · Iterative Simulator"
subtitle: "Paramecium, Didinium, and Lotka–Volterra"
format:
  html:
    theme: cosmo
    toc: true
    css: ../../styles.scss
execute:
  echo: true
  warning: false
  message: false
---

## Learning goals

1. Compare *Paramecium* alone vs with *Didinium*
2. Implement a discrete Lotka–Volterra step as a function
3. Store trajectories in vectors and iterate with a loop
4. Change one parameter and interpret biologically

## 1. Look at the data first

Real digitized series from **Gause (1934)** via `gauseR` (Fig. 21 monoculture; Fig. 32 with *Didinium*). See `data/README.md`.

```{r}
alone <- read.csv("data/paramecium_alone.csv")
both <- read.csv("data/paramecium_with_didinium.csv")

par(mfrow = c(1, 2))
plot(alone$day, alone$paramecium, type = "b", pch = 16,
     xlab = "Day", ylab = "Paramecium", main = "Alone (Fig. 21)")
plot(both$day, both$paramecium, type = "b", pch = 16,
     xlab = "Day", ylab = "Abundance", main = "With Didinium (Fig. 32)")
lines(both$day, both$didinium, type = "b", pch = 17, lty = 2)
imm <- both$day[both$immigration == "y"]
abline(v = imm, col = "grey70", lty = 3)
legend("topright", c("Prey", "Predator", "Immigration"),
       pch = c(16, 17, NA), lty = c(1, 2, 3), col = c("black", "black", "grey70"),
       bty = "n", cex = 0.8)
```

**Q1.** In 2–3 sentences: what is the biological contrast? Why might immigration days matter?

## 2. One-step update function

Fill in the missing pieces (or check the provided version):

```{r}
next_state <- function(N, P, r, a, b, m, dt) {
  dN <- (r * N - a * N * P) * dt
  dP <- (b * a * N * P - m * P) * dt
  c(N = max(N + dN, 0), P = max(P + dP, 0))
}

## try one step
next_state(N = 40, P = 2, r = 1.0, a = 0.1, b = 0.5, m = 0.6, dt = 0.05)
```

**Q2.** Did prey and predator increase or decrease? Does that match your verbal rules?

## 3. Vectors + loop

```{r}
T <- 200
dt <- 0.05
N <- numeric(T + 1)
P <- numeric(T + 1)
N[1] <- 40
P[1] <- 2

r <- 1.0
a <- 0.1
b <- 0.5
m <- 0.6

for (t in 1:T) {
  out <- next_state(N[t], P[t], r, a, b, m, dt)
  N[t + 1] <- out["N"]
  P[t + 1] <- out["P"]
}

time <- 0:T * dt
plot(time, N, type = "l", lwd = 2, xlab = "Time", ylab = "Abundance",
     main = "LV simulation")
lines(time, P, lwd = 2, lty = 2)
legend("topright", c("Prey", "Predator"), lty = 1:2, lwd = 2, bty = "n")
```

## 4. Experiment: change one parameter

Change **only** `a` (predation rate) by about ±25%. Re-run and compare.

**Q3.** What happened to prey? To predators? Link back to the Didinium treatment.

## 5. Optional: “no predator” run

Set `P[1] <- 0` (and keep it zero). What qualitative pattern do you recover from the alone-culture data?

## Wrap-up

- Main quantitative result:
- Main biological interpretation:
