---
title: "Week 1 Notebook - Predator-Prey Model"
author: "Quantitative Biology I"
format:
  html:
    toc: true
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

## Learning goals

By the end of this notebook, you should be able to:

1. Run a simple predator-prey simulation.
2. Describe how parameters influence population trajectories.
3. Compare baseline and modified ecological scenarios.

## Biological background

Predator-prey interactions are a classic example of ecological feedback:

- prey availability supports predator growth,
- predators reduce prey abundance,
- both populations can oscillate over time.

We implement a discrete-time Euler approximation of a Lotka-Volterra style model.

## 1. Setup

```{r}
# No extra package needed for this introductory notebook
```

## 2. Simulation function

```{r}
simulate_lv <- function(
  n_steps = 400,
  dt = 0.05,
  N0 = 40,    # initial prey
  P0 = 9,     # initial predator
  r = 1.0,    # prey growth
  a = 0.08,   # predation rate
  b = 0.06,   # conversion efficiency
  m = 0.7     # predator mortality
) {
  N <- numeric(n_steps + 1)
  P <- numeric(n_steps + 1)
  time <- seq(0, n_steps) * dt

  N[1] <- N0
  P[1] <- P0

  for (i in 1:n_steps) {
    dN <- r * N[i] - a * N[i] * P[i]
    dP <- b * a * N[i] * P[i] - m * P[i]

    N[i + 1] <- max(0, N[i] + dt * dN)
    P[i + 1] <- max(0, P[i] + dt * dP)
  }

  data.frame(time = time, prey = N, predator = P)
}
```

## 3. Baseline run

```{r}
baseline <- simulate_lv()
head(baseline)
```

```{r}
plot(
  baseline$time, baseline$prey,
  type = "l", lwd = 2, col = "darkgreen",
  xlab = "Time", ylab = "Abundance",
  main = "Predator-Prey Dynamics (Baseline)"
)
lines(baseline$time, baseline$predator, lwd = 2, col = "firebrick")
legend(
  "topright",
  legend = c("Prey", "Predator"),
  col = c("darkgreen", "firebrick"),
  lwd = 2, bty = "n"
)
```

## 4. Parameter exploration

Change one parameter and compare results to baseline.

### Example: increased predation rate

```{r}
high_predation <- simulate_lv(a = 0.10)

par(mfrow = c(1, 2))

plot(
  baseline$time, baseline$prey, type = "l", lwd = 2, col = "darkgreen",
  xlab = "Time", ylab = "Abundance", main = "Baseline"
)
lines(baseline$time, baseline$predator, lwd = 2, col = "firebrick")

plot(
  high_predation$time, high_predation$prey, type = "l", lwd = 2, col = "darkgreen",
  xlab = "Time", ylab = "Abundance", main = "Higher predation (a = 0.10)"
)
lines(high_predation$time, high_predation$predator, lwd = 2, col = "firebrick")

par(mfrow = c(1, 1))
```

## 5. Guided questions

Answer briefly in complete sentences:

1. How does increasing `a` affect prey peaks and troughs?
2. How does predator timing change relative to prey?
3. Which parameter do you predict most strongly controls predator persistence: `a`, `b`, or `m`? Why?

## 6. Mini challenge

Create and test one biologically motivated scenario:

- Example A: lower predator mortality (`m`)
- Example B: faster prey growth (`r`)
- Example C: lower conversion efficiency (`b`)

Include:

- one comparison plot,
- 3-4 bullet points interpreting the result.

## 7. Wrap-up

This model demonstrates how simple interaction rules can generate rich ecological dynamics.  
Next week, we extend this idea to competition between species.
