---
title: "Wright–Fisher Model — Mutation–Selection Balance"
subtitle: "Deleterious allele at a single locus"
author: "Quantitative Biology I"
format:
  html:
    toc: true
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

## Biological question

When a **deleterious mutation** keeps appearing in a population, why doesn't it disappear entirely — and why doesn't it take over?

At equilibrium, **new mutations** (rate $\mu$) are balanced by **selection removing them** (strength $s$, dominance $h$).

This notebook compares:

1. **Deterministic mutation–selection balance approximations**
2. **Stochastic Wright–Fisher simulations** (finite population size $N$)

## Model setup

Diploid locus with two alleles:

| Genotype | Fitness |
|----------|---------|
| $AA$ (wild type) | $1$ |
| $Aa$ | $1 - hs$ |
| $aa$ | $1 - s$ |

- $s$: selection coefficient against homozygous mutant ($s > 0$)
- $h$: dominance ($h = 0$ recessive, $h = 0.5$ additive, $h = 1$ dominant)
- $\mu$: forward mutation rate $A \to a$ per allele per generation

Each generation:

1. Selection on genotype frequencies (Hardy–Weinberg before selection)
2. Mutation
3. Binomial sampling of $2N$ alleles (Wright–Fisher drift)

## 1. Helper functions

```{r}
# Shared simulation settings (increase for even higher precision)
SIM_GENERATIONS <- 10000
SIM_BURNIN <- 3000
SIM_REPS_STANDARD <- 80
SIM_REPS_GRID <- 50

# Mutation-selection balance approximations (deleterious allele frequency q)
equilibrium_q <- function(mu, s, h, tol = 1e-10) {
  if (s <= 0) stop("s must be positive for a deleterious allele.")
  if (h > tol) {
    return(mu / (h * s))
  }
  sqrt(mu / s)
}

# Deterministic one-generation update (selection + mutation, no drift)
wf_generation_deterministic <- function(p, mu, s, h) {
  p <- max(0, min(1, p))
  w_AA <- 1
  w_Aa <- 1 - h * s
  w_aa <- 1 - s

  freq_AA <- (1 - p)^2
  freq_Aa <- 2 * p * (1 - p)
  freq_aa <- p^2

  w_bar <- freq_AA * w_AA + freq_Aa * w_Aa + freq_aa * w_aa
  if (w_bar <= 0) return(0)

  p_sel <- (freq_Aa * w_Aa + 2 * freq_aa * w_aa) / (2 * w_bar)
  p_sel <- max(0, min(1, p_sel))

  p_sel * (1 - mu) + (1 - p_sel) * mu
}

# Iterative deterministic equilibrium (infinite N, exact for this model)
equilibrium_q_deterministic <- function(mu, s, h, tol = 1e-12, max_iter = 5000) {
  p <- max(mu, equilibrium_q(mu, s, h))
  for (i in seq_len(max_iter)) {
    p_new <- wf_generation_deterministic(p, mu, s, h)
    if (abs(p_new - p) < tol) return(p_new)
    p <- p_new
  }
  p
}

# One Wright-Fisher generation (with drift)
wf_generation <- function(p, N, mu, s, h) {
  p <- max(0, min(1, p))
  w_AA <- 1
  w_Aa <- 1 - h * s
  w_aa <- 1 - s

  freq_AA <- (1 - p)^2
  freq_Aa <- 2 * p * (1 - p)
  freq_aa <- p^2

  w_bar <- freq_AA * w_AA + freq_Aa * w_Aa + freq_aa * w_aa
  if (w_bar <= 0) return(0)

  p_sel <- (freq_Aa * w_Aa + 2 * freq_aa * w_aa) / (2 * w_bar)
  p_sel <- max(0, min(1, p_sel))

  p_mut <- p_sel * (1 - mu) + (1 - p_sel) * mu
  p_mut <- max(0, min(1, p_mut))

  n_mut_alleles <- rbinom(1, size = 2 * N, prob = p_mut)
  n_mut_alleles / (2 * N)
}

# Simulate replicate trajectories; equilibrium = mean of per-replicate post-burnin means
simulate_wf <- function(
  N = 1000,
  mu = 1e-5,
  s = 0.05,
  h = 0.5,
  generations = SIM_GENERATIONS,
  p0 = NULL,
  n_reps = SIM_REPS_STANDARD,
  burnin = SIM_BURNIN
) {
  if (is.null(p0)) p0 <- equilibrium_q(mu, s, h)

  rep_means <- replicate(n_reps, {
    p <- p0
    post <- numeric(generations - burnin)
    for (g in seq_len(generations)) {
      p <- wf_generation(p, N, mu, s, h)
      if (g > burnin) post[g - burnin] <- p
    }
    mean(post)
  })

  list(
    rep_means = rep_means,
    mean_q = mean(rep_means),
    se_q = sd(rep_means) / sqrt(n_reps),
    sd_q = sd(rep_means),
    median_q = median(rep_means)
  )
}

# Simulate one (s, h) cell for heatmaps
simulate_grid_cell <- function(s, h, mu, N, n_reps, generations, burnin) {
  q_th <- equilibrium_q(mu, s, h)
  q_det <- equilibrium_q_deterministic(mu, s, h)
  sim <- simulate_wf(
    N = N, mu = mu, s = s, h = h,
    n_reps = n_reps, generations = generations, burnin = burnin
  )
  c(
    q_sim = sim$mean_q,
    q_se = sim$se_q,
    q_theory = q_th,
    q_det = q_det,
    ratio_approx = sim$mean_q / q_th,
    ratio_det = sim$mean_q / q_det
  )
}

# Optional: store full trajectories for visualization
simulate_wf_trajectories <- function(
  N = 1000, mu = 1e-5, s = 0.05, h = 0.5,
  generations = 6000, n_reps = 25, burnin = 1500, p0 = NULL
) {
  if (is.null(p0)) p0 <- equilibrium_q(mu, s, h)
  reps <- replicate(n_reps, {
    p <- p0
    traj <- numeric(generations)
    for (g in seq_len(generations)) {
      p <- wf_generation(p, N, mu, s, h)
      traj[g] <- p
    }
    traj
  })
  dim(reps) <- c(generations, n_reps)
  reps
}
```

## 2. Baseline simulation

```{r}
set.seed(42)

mu <- 1e-5
s <- 0.05
h <- 0.5
N <- 1000

q_approx <- equilibrium_q(mu, s, h)
q_det <- equilibrium_q_deterministic(mu, s, h)
sim <- simulate_wf(N = N, mu = mu, s = s, h = h, n_reps = SIM_REPS_STANDARD)
traj <- simulate_wf_trajectories(N = N, mu = mu, s = s, h = h, n_reps = 25)

cat("Approximate q:              ", signif(q_approx, 5), "\n")
cat("Deterministic equilibrium:", signif(q_det, 5), "\n")
cat("Simulated mean q (+/- SE):  ", signif(sim$mean_q, 5), "+/-", signif(sim$se_q, 5), "\n")
cat("Simulated median q:         ", signif(sim$median_q, 5), "\n")
```

```{r}
#| fig-width: 8
#| fig-height: 4
gens <- seq_len(nrow(traj))
matplot(
  gens, traj,
  type = "l", lty = 1, col = adjustcolor("steelblue", 0.35),
  xlab = "Generation", ylab = "Deleterious allele frequency (q)",
  main = sprintf("Wright-Fisher trajectories (N=%d, s=%.3f, h=%.2f)", N, s, h)
)
abline(h = q_approx, col = "firebrick", lwd = 2, lty = 2)
abline(h = q_det, col = "darkgreen", lwd = 2, lty = 3)
legend(
  "topright",
  legend = c("Replicate simulations", "Analytic approx.", "Deterministic eq."),
  col = c(adjustcolor("steelblue", 0.8), "firebrick", "darkgreen"),
  lty = c(1, 2, 3), lwd = c(1, 2, 2), bty = "n", cex = 0.85
)
```

## 3. Compare theory vs simulation across $s$ values

Fix $\mu$, $h$, and $N$; vary selection strength $s$ on a log-spaced grid.

```{r}
#| fig-width: 8
#| fig-height: 5
set.seed(1)

mu <- 1e-5
h <- 0.5
N <- 1000
s_values <- exp(seq(log(0.003), log(0.25), length.out = 16))

results_s <- data.frame(
  s = s_values,
  q_approx = NA_real_,
  q_det = NA_real_,
  q_sim_mean = NA_real_,
  q_sim_se = NA_real_
)

for (i in seq_along(s_values)) {
  sim_i <- simulate_wf(N = N, mu = mu, s = s_values[i], h = h, n_reps = SIM_REPS_STANDARD)
  results_s[i, ] <- list(
    s = s_values[i],
    q_approx = equilibrium_q(mu, s_values[i], h),
    q_det = equilibrium_q_deterministic(mu, s_values[i], h),
    q_sim_mean = sim_i$mean_q,
    q_sim_se = sim_i$se_q
  )
}

results_s
```

```{r}
#| fig-width: 8
#| fig-height: 4.5
ylim_s <- range(c(
  results_s$q_approx, results_s$q_det,
  results_s$q_sim_mean + 1.96 * results_s$q_sim_se,
  results_s$q_sim_mean - 1.96 * results_s$q_sim_se
))

plot(
  results_s$s, results_s$q_approx,
  log = "x", type = "l", lwd = 2, col = "firebrick",
  xlab = "Selection coefficient (s)",
  ylab = "Equilibrium deleterious allele frequency (q)",
  main = sprintf("Effect of s (mu=%g, h=%.1f, N=%d)", mu, h, N),
  ylim = ylim_s
)
lines(results_s$s, results_s$q_det, lwd = 2, col = "darkgreen", lty = 3)
points(results_s$s, results_s$q_sim_mean, pch = 16, col = "steelblue", cex = 0.9)
arrows(
  results_s$s, results_s$q_sim_mean - 1.96 * results_s$q_sim_se,
  results_s$s, results_s$q_sim_mean + 1.96 * results_s$q_sim_se,
  length = 0.03, angle = 90, col = "steelblue"
)
legend(
  "topright",
  legend = c("Analytic approx.", "Deterministic eq.", "Simulation mean +/- 95% CI"),
  col = c("firebrick", "darkgreen", "steelblue"),
  lty = c(1, 3, NA), pch = c(NA, NA, 16), lwd = c(2, 2, NA), bty = "n", cex = 0.85
)
```

**Interpretation:** stronger selection ($s \uparrow$) lowers equilibrium $q$, because deleterious alleles are removed faster. The approximation $q \approx \mu / (hs)$ tracks the simulation when $hs \gg \mu$.

## 4. Compare theory vs simulation across $h$ values

Fix $\mu$, $s$, and $N$; vary dominance $h$ on a fine grid (including near-recessive values).

```{r}
#| fig-width: 8
#| fig-height: 5
set.seed(2)

mu <- 1e-5
s <- 0.05
N <- 1000
h_values <- c(0, seq(0.02, 1, length.out = 15))

results_h <- data.frame(
  h = h_values,
  q_approx = NA_real_,
  q_det = NA_real_,
  q_sim_mean = NA_real_,
  q_sim_se = NA_real_
)

for (i in seq_along(h_values)) {
  sim_i <- simulate_wf(N = N, mu = mu, s = s, h = h_values[i], n_reps = SIM_REPS_STANDARD)
  results_h[i, ] <- list(
    h = h_values[i],
    q_approx = equilibrium_q(mu, s, h_values[i]),
    q_det = equilibrium_q_deterministic(mu, s, h_values[i]),
    q_sim_mean = sim_i$mean_q,
    q_sim_se = sim_i$se_q
  )
}

results_h
```

```{r}
#| fig-width: 8
#| fig-height: 4.5
ylim_h <- range(c(
  results_h$q_approx, results_h$q_det,
  results_h$q_sim_mean + 1.96 * results_h$q_sim_se,
  results_h$q_sim_mean - 1.96 * results_h$q_sim_se
))

plot(
  results_h$h, results_h$q_approx,
  type = "l", lwd = 2, col = "firebrick",
  xlab = "Dominance coefficient (h)",
  ylab = "Equilibrium deleterious allele frequency (q)",
  main = sprintf("Effect of h (mu=%g, s=%.2f, N=%d)", mu, s, N),
  ylim = ylim_h
)
lines(results_h$h, results_h$q_det, lwd = 2, col = "darkgreen", lty = 3)
points(results_h$h, results_h$q_sim_mean, pch = 16, col = "steelblue", cex = 0.9)
arrows(
  results_h$h, results_h$q_sim_mean - 1.96 * results_h$q_sim_se,
  results_h$h, results_h$q_sim_mean + 1.96 * results_h$q_sim_se,
  length = 0.03, angle = 90, col = "steelblue"
)
legend(
  "topright",
  legend = c("Analytic approx.", "Deterministic eq.", "Simulation mean +/- 95% CI"),
  col = c("firebrick", "darkgreen", "steelblue"),
  lty = c(1, 3, NA), pch = c(NA, NA, 16), lwd = c(2, 2, NA), bty = "n", cex = 0.85
)
```

**Interpretation:** when $h = 0$ (recessive deleterious allele), the square-root approximation $q \approx \sqrt{\mu / s}$ applies. When $h > 0$, the linear approximation $q \approx \mu / (hs)$ is more appropriate.

## 5. Heatmap: theory vs simulation over $(s, h)$ grid

High-resolution grid ($28 \times 28$) with parallel simulation across parameter combinations. Ratios are computed against both the **analytic approximation** and the **deterministic equilibrium** (infinite $N$).

```{r}
#| fig-width: 8
#| fig-height: 6
#| cache: true
set.seed(3)

mu <- 1e-5
N <- 1000

n_s <- 28
n_h <- 28
s_grid <- exp(seq(log(0.003), log(0.35), length.out = n_s))
h_grid <- seq(0, 1, length.out = n_h)

grid_idx <- expand.grid(i = seq_len(n_s), j = seq_len(n_h))

run_parallel <- function(x, FUN) {
  if (.Platform$OS.type == "unix") {
    parallel::mclapply(x, FUN, mc.cores = max(1, parallel::detectCores() - 1))
  } else {
    lapply(x, FUN)
  }
}

grid_results <- run_parallel(seq_len(nrow(grid_idx)), function(k) {
  i <- grid_idx$i[k]
  j <- grid_idx$j[k]
  simulate_grid_cell(
    s = s_grid[i], h = h_grid[j], mu = mu, N = N,
    n_reps = SIM_REPS_GRID,
    generations = SIM_GENERATIONS,
    burnin = SIM_BURNIN
  )
})

grid_mat <- do.call(rbind, grid_results)

ratio_approx <- matrix(grid_mat[, "ratio_approx"], nrow = n_s, ncol = n_h)
ratio_det <- matrix(grid_mat[, "ratio_det"], nrow = n_s, ncol = n_h)
q_sim_mat <- matrix(grid_mat[, "q_sim"], nrow = n_s, ncol = n_h)
q_se_mat <- matrix(grid_mat[, "q_se"], nrow = n_s, ncol = n_h)

data.frame(
  n_cells = nrow(grid_mat),
  ratio_approx_median = median(ratio_approx),
  ratio_det_median = median(ratio_det),
  frac_within_5pct_det = mean(abs(ratio_det - 1) <= 0.05),
  frac_within_10pct_det = mean(abs(ratio_det - 1) <= 0.10),
  max_abs_error_q = max(abs(q_sim_mat - matrix(grid_mat[, "q_det"], nrow = n_s, ncol = n_h)))
)
```

```{r}
#| fig-width: 9
#| fig-height: 6.5
filled.contour(
  x = h_grid,
  y = s_grid,
  z = ratio_det,
  xlab = "Dominance coefficient (h)",
  ylab = "Selection coefficient (s)",
  main = "Simulation / deterministic equilibrium",
  color.palette = function(n) hcl.colors(n, "RdYlBu", rev = TRUE),
  plot.axes = {
    axis(1)
    axis(2)
    contour(h_grid, s_grid, ratio_det, levels = 1, add = TRUE, lwd = 2, col = "black")
  },
  key.axes = axis(4, col = "black", cex.axis = 0.85),
  plot.title = title(main = "Simulation / deterministic equilibrium (1 = perfect match)", cex.main = 1.05)
)
mtext("Color: simulated q / deterministic q", side = 4, line = 2.5, cex = 0.85)
```

```{r}
#| fig-width: 9
#| fig-height: 6.5
filled.contour(
  x = h_grid,
  y = s_grid,
  z = ratio_approx,
  xlab = "Dominance coefficient (h)",
  ylab = "Selection coefficient (s)",
  main = "Simulation / analytic approximation",
  color.palette = function(n) hcl.colors(n, "RdYlBu", rev = TRUE),
  plot.axes = {
    axis(1)
    axis(2)
    contour(h_grid, s_grid, ratio_approx, levels = 1, add = TRUE, lwd = 2, col = "black")
  },
  key.axes = axis(4, col = "black", cex.axis = 0.85),
  plot.title = title(main = "Simulation / analytic approximation (1 = perfect match)", cex.main = 1.05)
)
mtext("Color: simulated q / approx. q", side = 4, line = 2.5, cex = 0.85)
```

```{r}
#| fig-width: 9
#| fig-height: 6.5
filled.contour(
  x = h_grid,
  y = s_grid,
  z = q_sim_mat,
  xlab = "Dominance coefficient (h)",
  ylab = "Selection coefficient (s)",
  main = "Mean simulated equilibrium allele frequency",
  color.palette = function(n) hcl.colors(n, "Blues"),
  plot.axes = {
    axis(1)
    axis(2)
  },
  key.axes = axis(4, col = "black", cex.axis = 0.85),
  plot.title = title(main = "Mean simulated equilibrium q", cex.main = 1.05)
)
mtext("Color: mean simulated q", side = 4, line = 2.5, cex = 0.85)
```

```{r}
#| fig-width: 9
#| fig-height: 6.5
# Simulation uncertainty across the grid (SE of replicate means)
filled.contour(
  x = h_grid,
  y = s_grid,
  z = q_se_mat,
  xlab = "Dominance coefficient (h)",
  ylab = "Selection coefficient (s)",
  main = "Standard error of simulated equilibrium",
  color.palette = function(n) hcl.colors(n, "Oranges"),
  plot.axes = {
    axis(1)
    axis(2)
  },
  key.axes = axis(4, col = "black", cex.axis = 0.85),
  plot.title = title(main = "SE of simulated q (precision of estimate)", cex.main = 1.05)
)
mtext("Color: SE across replicate means", side = 4, line = 2.5, cex = 0.85)
```

Ratios near **1** mean the approximation works well. Deviations often occur when:

- $q$ is not small (weak selection / high $\mu$)
- $N$ is small (drift dominates)
- $h \approx 0$ (boundary between approximations)

## 6. Effect of population size

```{r}
#| fig-width: 8
#| fig-height: 4.5
set.seed(4)

mu <- 1e-5
s <- 0.05
h <- 0.5
N_values <- unique(round(exp(seq(log(100), log(20000), length.out = 12))))

q_approx <- equilibrium_q(mu, s, h)
q_det <- equilibrium_q_deterministic(mu, s, h)

results_N <- data.frame(
  N = N_values,
  q_sim_mean = NA_real_,
  q_sim_se = NA_real_
)

for (i in seq_along(N_values)) {
  sim_i <- simulate_wf(N = N_values[i], mu = mu, s = s, h = h, n_reps = SIM_REPS_STANDARD)
  results_N[i, ] <- list(N = N_values[i], q_sim_mean = sim_i$mean_q, q_sim_se = sim_i$se_q)
}

plot(
  results_N$N, results_N$q_sim_mean, log = "x", pch = 16, col = "steelblue",
  xlab = "Population size (N)", ylab = "Mean simulated q",
  main = "Finite-N effects on mutation-selection balance",
  ylim = range(c(q_approx, q_det, results_N$q_sim_mean + 1.96 * results_N$q_sim_se))
)
arrows(
  results_N$N, results_N$q_sim_mean - 1.96 * results_N$q_sim_se,
  results_N$N, results_N$q_sim_mean + 1.96 * results_N$q_sim_se,
  length = 0.03, angle = 90, col = "steelblue"
)
abline(h = q_approx, col = "firebrick", lty = 2, lwd = 2)
abline(h = q_det, col = "darkgreen", lty = 3, lwd = 2)
legend(
  "topright",
  legend = c("Simulation +/- 95% CI", "Analytic approx.", "Deterministic eq."),
  col = c("steelblue", "firebrick", "darkgreen"),
  pch = c(16, NA, NA), lty = c(NA, 2, 3), lwd = c(NA, 2, 2), bty = "n", cex = 0.85
)
```

## 7. Guided questions

1. For fixed $\mu$, explain why equilibrium $q$ decreases as $s$ increases.
2. Why does recessive deleterious variation ($h = 0$) reach higher equilibrium frequency than dominant deleterious variation ($h = 1$) at the same $s$?
3. Where do simulations deviate most from the deterministic approximation? Is drift or breakdown of the approximation the main reason?
4. How would you expect the results to change if you added backward mutation ($a \to A$)?

## Reference approximations used

| Case | Equilibrium frequency of deleterious allele |
|------|---------------------------------------------|
| $h > 0$ | $\hat{q} \approx \dfrac{\mu}{h s}$ |
| $h = 0$ (recessive) | $\hat{q} \approx \sqrt{\dfrac{\mu}{s}}$ |

These are **approximations**. The notebook also computes a **deterministic equilibrium** by iterating selection + mutation without drift (infinite $N$), which is the sharper theoretical reference for comparing simulations at finite $N$.

**Simulation settings in this notebook:** `10,000` generations, `3,000` burn-in, `80` replicates (standard analyses), `50` replicates per heatmap cell, `28 x 28` parameter grid.
